---
title: "changepoint analysis"
author: "Afiqah"
date: "2023-02-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workflow

## 1. Prepare R
### 1.1 Set working directory
### 1.2 Load packages
### 1.3 Load data

## 2. Analysis
### 2.1 Explore data
### 2.2 Prepare dataframe for variables
### 2.3 Changepoint analysis
#### 2.3.1 Overall trend
#### 2.3.2 Triage category
#### 2.3.3 Discharge outcome
#### 2.3.4 ED discharge diagnosis

# Analysis

## 1. Prepare R
### 1.1 Set working directory
```{r}
getwd()
```

### 1.2 Load packages
```{r}
library(dplyr)
library(ggplot2)
library(gtsummary)
library(EnvCpt)
library(changepoint)
library(changepoint.influence)
library(ggpubr)
```

### 1.3 Load data
```{r}
data <- read.csv('EDbyWeek.csv')
```

## 2. Analysis
### 2.1 Explore data
```{r}
summary(data)
```

```{r}
ggplot(data, aes(x=EpidWeek, y=total)) + geom_line()
```
### 2.2 Prepare dataframe for variables
```{r}
all <- c(data$total)
green <- c(data$green)
yellow <- c(data$yellow)
red <- c(data$red)
discharge <- c(data$discharge)
admission <- c(data$admission)
death <- c(data$death)
allergic <- c(data$allergic)
assault <- c(data$assault)
trauma <- c(data$traumaburn)
cns <- c(data$cns)
derm <- c(data$derm)
ent <- c(data$ent)
cd <- c(data$cd)
gi <- c(data$gi)
msp <- c(data$msp)
neo <- c(data$neo)
nephro <- c(data$nephro)
ong <- c(data$ong)
omfs <- c(data$omfs)
ophthal <- c(data$ophthal)
others <- c(data$others)
poison <- c(data$poison)
psy <- c(data$psy)
respi <- c(data$respi)
```

### 2.3 Changepoint analysis

#### 2.3.1 Overall trend
```{r}

### Primary model

cpt.all <- envcpt(all)

BIC(cpt.all) # returns the BIC for each model.
which.min(BIC(cpt.all)) 
plot(cpt.all,type='fit') # plots the fits
plot(cpt.all,type="bic") # plots the bic values

### Selection of model based on best BIC

cpt.all$meanar2cpt

cpts(cpt.all$meanar2cpt)


### Plot model

plot(cpt.all[[which.min(BIC(cpt.all))+1]])
abline(v=cpts(cpt.all$meanar2cpt), col='blue')

### Test for model fitness
cpt.all.fit = envcpt(all,models=c("meanar2cpt"))$meanar2cpt
resid.all = all - rep(param.est(cpt.all.fit)$beta[,1],times=seg.len(cpt.all.fit)) -
  rep(param.est(cpt.all.fit)$beta[,2],times=seg.len(cpt.all.fit))*1:length(all)
corrected.all = resid.all + param.est(cpt.all.fit)$beta[3,1] + param.est(cpt.all.fit)$beta[3,2]*1:length(all)

shapiro.test(corrected.all)

ks.test(resid.all, pnorm, mean = mean(corrected.all), sd = sd(corrected.all))

qqnorm(corrected.all)
qqline(corrected.all)
```
#### 2.3.2 Triage category

##### 2.3.2.1 Triage category: Green
```{r}

### Primary model

cpt.green <- envcpt(green)

BIC(cpt.green) # returns the BIC for each model.
which.min(BIC(cpt.green)) 
plot(cpt.green,type='fit') # plots the fits
plot(cpt.green,type="bic") # plots the bic values

### Selection of model based on best BIC

cpt.green$meanar2cpt

cpts(cpt.green$meanar2cpt)


### Plot model

plot(cpt.green[[which.min(BIC(cpt.green))+1]])
abline(v=cpts(cpt.green$meanar2cpt), col='blue')

### Test for model fitness
cpt.green.fit = envcpt(all,models=c("meanar2cpt"))$meanar2cpt
resid.green = green - rep(param.est(cpt.green.fit)$beta[,1],times=seg.len(cpt.green.fit)) -
  rep(param.est(cpt.green.fit)$beta[,2],times=seg.len(cpt.green.fit))*1:length(green)
corrected.green = resid.green + param.est(cpt.green.fit)$beta[3,1] + param.est(cpt.green.fit)$beta[3,2]*1:length(green)

shapiro.test(corrected.green)

ks.test(resid.green, pnorm, mean = mean(corrected.green), sd = sd(corrected.green))

qqnorm(corrected.green)
qqline(corrected.green)
```

##### 2.3.2.2 Triage category: Yellow
```{r}

### Primary model

cpt.yellow <- envcpt(yellow)

BIC(cpt.yellow) # returns the BIC for each model.
which.min(BIC(cpt.yellow)) 
plot(cpt.yellow,type='fit') # plots the fits
plot(cpt.yellow,type="bic") # plots the bic values

### Selection of model based on best BIC

cpt.yellow$meanar1cpt

cpts(cpt.yellow$meanar1cpt)


### Plot model

plot(cpt.yellow[[which.min(BIC(cpt.yellow))+1]])
abline(v=cpts(cpt.yellow$meanar1cpt), col='blue')

### Test for model fitness
cpt.yellow.fit = envcpt(yellow,models=c("meanar1cpt"))$meanar1cpt
resid.yellow = yellow - rep(param.est(cpt.yellow.fit)$beta[,1],times=seg.len(cpt.yellow.fit)) -
  rep(param.est(cpt.yellow.fit)$beta[,2],times=seg.len(cpt.yellow.fit))*1:length(yellow)
corrected.yellow = resid.yellow + param.est(cpt.yellow.fit)$beta[3,1] + param.est(cpt.yellow.fit)$beta[3,2]*1:length(yellow)

shapiro.test(corrected.yellow)

ks.test(resid.yellow, pnorm, mean = mean(corrected.yellow), sd = sd(corrected.yellow))

qqnorm(corrected.yellow)
qqline(corrected.yellow)
```

#### 2.3.2.3 Triage category: Red
```{r}

### Primary model

cpt.red <- envcpt(red)

BIC(cpt.red) 
which.min(BIC(cpt.red)) 
plot(cpt.red,type='fit') 
plot(cpt.red,type="bic")

### Selection of model based on best BIC

cpt.red$meanar1cpt

cpts(cpt.red$meanar1cpt)


### Plot model

plot(cpt.red[[which.min(BIC(cpt.red))+1]]) + abline(v=cpts(cpt.red$meanar1cpt), col='blue')

### Test for model fitness
cpt.red.fit = envcpt(red,models=c("meanar1cpt"))$meanar1cpt
resid.red = red - rep(param.est(cpt.red.fit)$beta[,1],times=seg.len(cpt.red.fit)) -
  rep(param.est(cpt.red.fit)$beta[,2],times=seg.len(cpt.red.fit))*1:length(red)
corrected.red = resid.red + param.est(cpt.red.fit)$beta[3,2] + param.est(cpt.red.fit)$beta[3,1]*1:length(red)

shapiro.test(corrected.red)

ks.test(resid.red, pnorm, mean = mean(corrected.red), sd = sd(corrected.red))

qqnorm(corrected.red)
qqline(corrected.red)
```
#### 2.3.3 Visit outcome

##### 2.3.3.1 Visit outcome: Discharge
```{r}

### Primary model

cpt.discharge <- envcpt(discharge)

BIC(cpt.discharge) 
which.min(BIC(cpt.discharge)) 
plot(cpt.discharge,type='fit') 
plot(cpt.discharge,type="bic")

### Selection of model based on best BIC

cpt.discharge$meanar2cpt

cpts(cpt.discharge$meanar2cpt)


### Plot model

plot.discharge <- plot(cpt.discharge[[which.min(BIC(cpt.discharge))+1]]) + abline(v=cpts(cpt.discharge$meanar2cpt), col='blue')

### Test for model fitness
cpt.discharge.fit = envcpt(discharge,models=c("meanar2cpt"))$meanar2cpt
resid.discharge = discharge - rep(param.est(cpt.discharge.fit)$beta[,1],times=seg.len(cpt.discharge.fit)) -  rep(param.est(cpt.discharge.fit)$beta[,2],times=seg.len(cpt.discharge.fit))*1:length(discharge)
corrected.discharge = resid.discharge + param.est(cpt.discharge.fit)$beta[3,2] + param.est(cpt.discharge.fit)$beta[3,1]*1:length(discharge)

shapiro.test(corrected.discharge)

ks.test(resid.discharge, pnorm, mean = mean(corrected.discharge), sd = sd(corrected.discharge))

qqnorm(corrected.discharge)
qqline(corrected.discharge)
```

##### 2.3.3.2 Visit outcome: Admission
```{r}

### Primary model

cpt.admission <- envcpt(admission)

BIC(cpt.admission) 
which.min(BIC(cpt.admission)) 
plot(cpt.admission,type='fit') 
plot(cpt.admission,type="bic")

### Selection of model based on best BIC

cpt.admission$meanar2cpt

cpts(cpt.admission$meanar2cpt)


### Plot model

plot(cpt.admission[[which.min(BIC(cpt.admission))+1]]) + abline(v=cpts(cpt.admission$meanar2cpt), col='blue')

### Test for model fitness
cpt.admission.fit = envcpt(admission,models=c("meanar2cpt"))$meanar2cpt

resid.admission = admission - (rep(param.est(cpt.admission.fit)$beta[,1],times=seg.len(cpt.admission.fit))) -  (rep(param.est(cpt.admission.fit)$beta[,2],times=seg.len(cpt.admission.fit))*1:length(admission))

corrected.admission = resid.admission + param.est(cpt.admission.fit)$beta[2,2] + param.est(cpt.admission.fit)$beta[2,1]*1:length(admission)

shapiro.test(corrected.admission)

ks.test(resid.admission, pnorm, mean = mean(corrected.admission), sd = sd(corrected.admission))

qqnorm(corrected.admission)
qqline(corrected.admission)
```

##### 2.3.3.3 Visit outcome: Death
```{r}

### Primary model

cpt.death <- envcpt(death)

BIC(cpt.death) 
which.min(BIC(cpt.death)) 
plot(cpt.death,type='fit') 

### Selection of model based on best BIC

cpt.death$trendar1cpt

cpts(cpt.death$trendar1cpt)


### Plot model

plot(cpt.death$trendar1cpt)

### Test for model fitness
cpt.death.fit = envcpt(death,models=c("trendar1cpt"))$trendar1cpt

resid.death = death - (rep(param.est(cpt.death.fit)$beta[,1],times=seg.len(cpt.death.fit))) -  (rep(param.est(cpt.death.fit)$beta[,2],times=seg.len(cpt.death.fit))*1:length(death))

corrected.death = resid.death + param.est(cpt.death.fit)$beta[2,2] + param.est(cpt.death.fit)$beta[2,1]*1:length(death)

shapiro.test(corrected.admission)

ks.test(resid.admission, pnorm, mean = mean(corrected.admission), sd = sd(corrected.admission))

qqnorm(corrected.admission)
qqline(corrected.admission)
```

#### 2.3.4 ED discharge diagnosis

##### 2.3.4.1 Bump chart top 10 ED discharge diagnosis

```{r}
my_theme <- function() {

  # Colors
  color.background = "white"
  color.text = "#22211d"

  # Begin construction of chart
  theme_bw(base_size=15) +

    # Format background colors
    theme(panel.background = element_rect(fill=color.background, color=color.background)) +
    theme(plot.background  = element_rect(fill=color.background, color=color.background)) +
    theme(panel.border     = element_rect(color=color.background)) +
    theme(strip.background = element_rect(fill=color.background, color=color.background)) +

    # Format the grid
    theme(panel.grid.major.y = element_blank()) +
    theme(panel.grid.minor.y = element_blank()) +
    theme(axis.ticks       = element_blank()) +

    # Format the legend
    theme(legend.position = "none") +

    # Format title and axis labels
    theme(plot.title       = element_text(color=color.text, size=20, face = "bold")) +
    theme(axis.title.x     = element_text(size=14, color="black", face = "bold")) +
    theme(axis.title.y     = element_text(size=14, color="black", face = "bold", vjust=1.25)) +
    theme(axis.text.x      = element_text(size=10, vjust=0.5, hjust=0.5, color = color.text)) +
    theme(axis.text.y      = element_text(size=10, color = color.text)) +
    theme(strip.text       = element_text(face = "bold")) +

    # Plot margins
    theme(plot.margin = unit(c(0.35, 0.2, 0.3, 0.35), "cm"))
}
```


```{r}
rank <- read.csv('dxrank.csv')

rank$Period2 <- as.factor(rank$Period2)

df.rank <- rank %>% 
  group_by(Period) %>% 
  arrange(Period2, Rank, Diagnosis) %>% 
  mutate(ranking = row_number()) %>% 
  as.data.frame()

head(df.rank)

library(ggbump)

show.top.n <- 10

ggplot(data = df.rank, aes(x = Period2, y = ranking, group = Diagnosis)) +
  geom_line(aes(color = Diagnosis, alpha = 1), size = 2) +
  geom_point(aes(color = Diagnosis, alpha = 1), size = 3) +
  geom_point(color = "#FFFFFF", size = 1) +
  scale_y_reverse(breaks = 1:18) +
  coord_cartesian(ylim = c(1,18)) + 
  theme(legend.position = "none") +
  my_theme() 

```

##### 2.3.4.2 Diagnosis: Respi
```{r}
### Primary model

cpt.respi <- envcpt(respi)

BIC(cpt.respi) 
which.min(BIC(cpt.respi)) 
plot(cpt.respi,type="bic")

### Selection of model based on best BIC

cpt.respi$meanar2cpt

cpts(cpt.respi$meanar2cpt)
```

```{r}
plot(cpt.respi[[which.min(BIC(cpt.respi))+1]]) + abline(v=cpts(cpt.respi$meanar2cpt), col='blue')
```

##### 2.3.4.3 Diagnosis: CNS
```{r}
### Primary model

cpt.cns <- envcpt(cns)

BIC(cpt.cns) 
which.min(BIC(cpt.cns)) 
plot(cpt.cns,type="bic")

### Selection of model based on best BIC

plot(cpt.cns[[which.min(BIC(cpt.cns))+1]]) + abline(v=cpts(cpt.cns$meancpt), col='blue')

```

##### 2.3.4.4 Diagnosis: Fever and CD
```{r}
### Primary model

cpt.cd <- envcpt(cd)

BIC(cpt.cd) 
which.min(BIC(cpt.cd)) 
plot(cpt.cd,type="bic")

### Selection of model based on best BIC

cpt.cd$trendar2cpt

cpts(cpt.cd$trendar2cpt)

plot(cpt.cd[[which.min(BIC(cpt.cd))+1]]) + abline(v=cpts(cpt.cd$trendar2cpt), col='blue')
```

##### 2.3.4.5 Diagnosis: Burn or trauma
```{r}
### Primary model

cpt.trauma <- envcpt(trauma)

BIC(cpt.trauma) 
which.min(BIC(cpt.trauma)) 
plot(cpt.trauma,type="bic")

### Selection of model based on best BIC

cpt.trauma$trendar2

plot(cpt.trauma$trendar2cpt)
```

##### 2.3.4.6 Diagnosis: GI
```{r}
### Primary model

cpt.gi <- envcpt(gi)

BIC(cpt.gi) 
which.min(BIC(cpt.gi)) 
plot(cpt.gi,type="bic")

### Selection of model based on best BIC

cpt.gi$meanar2cpt

plot(cpt.gi[[which.min(BIC(cpt.gi))+1]]) + abline(v=cpts(cpt.gi$meanar2cpt), col='blue')

```

##### 2.3.4.7 Diagnosis: Neonatal
```{r}
### Primary model

cpt.neo <- envcpt(neo)

BIC(cpt.neo) 
which.min(BIC(cpt.neo)) 
plot(cpt.neo,type="bic")

### Selection of model based on best BIC

cpt.neo$trendar2cpt

plot(cpt.neo[[which.min(BIC(cpt.neo))+1]]) + abline(v=cpts(cpt.neo$trendar2cpt), col='blue')
```